---
title: "Zurich Public Pools App - Capstone Project"
author: "Kerim Lengwiler"
date: "2024-05-28"
output: 
  html_document:
  toc: true
toc_float: true
theme: united
---

# Zurich Public Pools App - Capstone Project

## Introduction
As summer approaches, many residents and visitors of Zurich are looking forward to enjoying the city's 18 public swimming pools. However, finding relevant information is like a paper chase on the city of Zurich's website and for visitors not familiar with the city, even finding the pools can be a challenge. Using the techniques learned in the course "Data Mining for the Social Sciences using R" at the University of Lucerne, this project aims to change that by creating a user-friendly and free of any cost web application  to access information about Zurich's public pools. Therefore, the projects goals are as follows:

* Create a Shiny app that provides a user-friendly interface to access and filter information (water temperature, opening status) about Zurich's public pools using the Baditicker-API provided by the city of Zurich.
* Provide directions from the user's location to any selected public pool using the Open Route Service API.


## Data, Methods and Access
### Data Sources
To produce the app, two open access data sources are used:

* The [Baditicker-API](https://www.stadt-zuerich.ch/stzh/bathdatadownload) is used to provide real-time information about the city of Zurich's public pools. 
* The [Open Route Service API](https://openrouteservice.org) is used to provide directions from the user's location to any selected public pool.

### Access
The web application can be accessed and tested through this [link](https://capstone-project-pools.shinyapps.io/scripts_app/).

### Methods
The app is built using the Shiny package in R based on data that is requested and then parsed from the Baditicker-API and from the Open Route Service API.

### Use of LLMs
While not a complete beginner using R, until this project, I have only used R for statistical data analysis and visualization. I therefore used github Copilot as well as ChatGPT 4 to help writing the basic code for the shiny app and to understand and solve issues that arrived with the use of the APIs and the gradual improvement of the application.


## R Code

## General Setup
### Libraries and API Access
```{r message=FALSE, warning=FALSE, fig.show='hide', results='hide'}
options(repos = c(CRAN = "https://cran.rstudio.com/"))

# Setup libraries
install.packages("remotes")
install.packages("openrouteservice")
install.packages("shiny")
install.packages("httr")
install.packages("XML")
install.packages("bslib")
install.packages("leaflet")
install.packages("tidyverse")
install.packages("sf")
install.packages("shinydashboard")
install.packages("fresh")
library(remotes)
library(openrouteservice)
library(shiny)
library(httr)
library(XML)
library(bslib)
library(leaflet)
library(tidyverse)
library(sf)
library(shinydashboard)
library(fresh)

ors_api_key <- Sys.getenv("ORS_API_KEY") # API Key for OpenRouteService

url <- "https://www.stadt-zuerich.ch/stzh/bathdatadownload"# API for pool data
```

### Fetching and Processing Pool Data
```{r message=FALSE, warning=FALSE, fig.show='hide', results='hide'}

# Make the request for pool data
pool_response <- GET(url = url, query = list()) 
http_status(pool_response)

xml_data <- xmlParse(rawToChar(pool_response$content))
pool <- getNodeSet(xml_data, "//bath")

# Extract data for each pool
pool_details <- lapply(pool, function(pool) { 
  title <- xmlValue(getNodeSet(pool, "./title")[[1]])
  temperatureWater <- xmlValue(getNodeSet(pool, "./temperatureWater")[[1]])
  poiid <- xmlValue(getNodeSet(pool, "./poiid")[[1]])
  dateModified <- xmlValue(getNodeSet(pool, "./dateModified")[[1]])
  openClosedTextPlain <- xmlValue(getNodeSet(pool, "./openClosedTextPlain")[[1]])
  urlPage <- xmlValue(getNodeSet(pool, "./urlPage")[[1]])
  pathPage <- xmlValue(getNodeSet(pool, "./pathPage")[[1]])
  
# Return a list of pool details
  list(Title = title, Wassertemperatur = temperatureWater, ID = poiid, 
       Update = dateModified, Status = openClosedTextPlain, 
       URL_Page = urlPage, Path_Page = pathPage)
})

# Convert the list to a data frame for easier manipulation and viewing
df_pools <- do.call(rbind.data.frame, pool_details)

if (class(df_pools$Wassertemperatur) %in% c("factor", "character")) {
  df_pools$Wassertemperatur <- as.numeric(as.character(df_pools$Wassertemperatur))
}

# Add Coordinates for the pools
df_pools_coordinates <- data.frame(
  Title = c("Flussbad Au-Höngg","Flussbad Oberer Letten","Flussbad Unterer Letten", "Flussbad Unterer Letten Flussteil", "Frauenbad Stadthausquai", "Freibad Allenmoos","Freibad Auhof",
            "Freibad Dolder", "Freibad Heuried", "Freibad Letzigraben", "Freibad Seebach", "Freibad Zwischen den Hölzern", "Hallenbad Bläsi", "Hallenbad Bungertwies", "Hallenbad City",
            "Hallenbad Leimbach", "Hallenbad Oerlikon", "Männerbad Schanzengraben ", "Seebad Enge", "Seebad Katzensee", "Seebad Utoquai", "Strandbad Mythenquai",
            "Strandbad Tiefenbrunnen","Strandbad Wollishofen ", "Wärmebad Käferberg" ),
  Latitude = c(47.39910, 47.38532, 47.39005, 47.38895, 47.36839, 47.40549, 47.40876, 47.37548, 47.36778, 47.37851, 47.42352, 47.40904, 47.40133, 47.37201, 47.37209, 47.32652,
               47.41023, 47.37122, 47.36170, 47.42844, 47.36176, 47.35451, 47.35239, 47.34112, 47.39942),
  Longitude = c(8.489321, 8.534956, 8.528608, 8.529391, 8.542030, 8.539013, 8.571462, 8.576269, 8.505399, 8.498366, 8.548195, 8.469622, 8.501876, 8.560206, 8.532868, 8.513683, 8.556734, 
                8.532703, 8.536701, 8.495608, 8.547013, 8.534713, 8.557461, 8.537336, 8.518000))


# Merging the data frames
df_pools <- merge(df_pools, df_pools_coordinates, by = "Title")
```

## Shiny App
### Setup
```{r message=FALSE, warning=FALSE, fig.show='hide', results='hide'}
# Creating Theme
capstone_theme <- create_theme(
  adminlte_color(
    light_blue = "#434C5E"
  ),
  adminlte_sidebar(
    width = "400px",
    dark_bg = "#D8DEE9",
    dark_hover_bg = "#81A1C1",
    dark_color = "#2E3440"
  ),
  adminlte_global(
    content_bg = "#FFF",
    box_bg = "#D8DEE9", 
    info_box_bg = "#D8DEE9"
  )
)

# Route Function (Directions) from Open Route Service API
get_route <- function(from_coords, to_coords, transport_type) {
  req(from_coords, to_coords, transport_type)
  tryCatch({
    route <- ors_directions(api_key = ors_api_key,
                            coordinates = list(from_coords, to_coords),
                            profile = transport_type, 
                            format = "geojson")
    print(paste("Route fetched from", from_coords, "to", to_coords, "using", transport_type))
    return(route)
  }, error = function(e) {
    print(paste("Error fetching route:", e$message))
    return(NULL)
  })
}
```

### User Interface
```{r message=FALSE, warning=FALSE, fig.show='hide', results='hide'}
# Setting up header of the dashboard: serves as main title of the application.
ui <- dashboardPage(
  dashboardHeader(title = tags$span("Freibäder Stadt Zürich", style = "font-weight: bold;")),
  
# Setting up the sidebar of the dashboard: serves as the main user interface for the application.
# Slider Input: Allows user to filter pools based on water temperature. It dynamically sets the minimum and maximum values based on the data available.
# Select Inputs: Include options to filter pools by status (open or closed), by name and to select transportation type (walking, cycling and driving).
# Action buttons: Provide functionalities to display routes, clear routes, and update user location.
  
  dashboardSidebar(
    width = 250, 
    sliderInput("temperatur", "Wassertemperatur", 
                min = min(df_pools$Wassertemperatur, na.rm = TRUE), 
                max = max(df_pools$Wassertemperatur, na.rm = TRUE), 
                value = c(min(df_pools$Wassertemperatur, na.rm = TRUE), 
                          max(df_pools$Wassertemperatur, na.rm = TRUE))),
    selectInput("status", "Status", choices = c("Alle", "offen", "geschlossen")),
    selectInput("title", "Name des Bades", choices = c("Alle", "Please select")),
    selectInput("transport_type", "Transportart", 
                choices = c("Fussweg" = "foot-walking", "Fahrrad" = "cycling-regular", "Auto" = "driving-car")),
    div(
      style = "display: flex; justify-content: flex-start; margin-top: 20px; margin-bottom: 20px;", 
      actionButton("show_route", "Route anzeigen", style = "margin-right: 5px;", icon = icon("route")),
      actionButton("clear_routes", "Route löschen", icon = icon("eraser"))
    ),
    actionButton("update_location", "Standort aktualisieren", icon = icon("location-arrow"))
  ),

# Setting up the body of the dashboard: serves as the main content area of the application.
# Leaflet Output: Displays the map with the pools and routes.
# Custom CSS: Adjusts the height of the map to fit the screen size.
# Java Script: Fetches the user's location and displays an error message if geolocation is not supported.
# Java Script: Sends the user's location to the server when the 'Standort aktualisieren' button is clicked.
# Java Script: Sends the route data to the server when the 'Route anzeigen' button is clicked.

  dashboardBody(
    use_theme(capstone_theme),
    tags$head(
      tags$style(HTML("
        .leaflet-container {
          height: calc(100vh - 80px) !important; 
        }
      ")),
      tags$script(HTML("
        document.addEventListener('DOMContentLoaded', function() {
          // Geolocation setup
          if (navigator.geolocation) {
            navigator.geolocation.getCurrentPosition(showPosition, showError);
          } else {
            Shiny.setInputValue('geolocation_error', 'Geolocation is not supported by this browser.');
          }

          function showPosition(position) {
            Shiny.setInputValue('user_lat', position.coords.latitude);
            Shiny.setInputValue('user_lon', position.coords.longitude);
            Shiny.setInputValue('geolocation', 'Lat: ' + position.coords.latitude + ', Lon: ' + position.coords.longitude);
          }

          function showError(error) {
            switch(error.code) {
              case error.PERMISSION_DENIED:
                Shiny.setInputValue('geolocation_error', 'User denied the request for Geolocation.');
                break;
              case error.POSITION_UNAVAILABLE:
                Shiny.setInputValue('geolocation_error', 'Location information is unavailable.');
                break;
              case error.TIMEOUT:
                Shiny.setInputValue('geolocation_error', 'The request to get user location timed out.');
                break;
              case error.UNKNOWN_ERROR:
                Shiny.setInputValue('geolocation_error', 'An unknown error occurred.');
                break;
            }
          }

          // Sidebar hiding functionality on 'Route anzeigen' button click
          $(document).on('shiny:inputchanged', function(event) {
            if (event.name === 'show_route') {
              $('#sidebarID').css('width', '0px');
              $('#sidebarID').css('visibility', 'hidden');
              $('.content-wrapper').css('margin-left', '0px');
            }
          });
        });
      "))
    ),
    leafletOutput("map", width = "100%", height = "100%")
  )
)
```

### Server Logic
```{r message=FALSE, warning=FALSE, fig.show='hide', results='hide'}
server <- function(input, output, session) {
  session$userData <- reactiveValues(selected_pool = NULL, user_location = NULL)
  
# Observe the map marker click event  
  observeEvent(input$map_marker_click, {
    req(input$map_marker_click)
    session$userData$selected_pool <- c(input$map_marker_click$lng, input$map_marker_click$lat)  
    showNotification("Standort Freibad ausgewählt.", type = "message")
  })
  
  observeEvent(input$show_route, {
    req(session$userData$selected_pool, input$user_lat, input$user_lon, input$transport_type)
    session$userData$user_location <- c(input$user_lon, input$user_lat)  
    
    from_coords <- session$userData$user_location
    to_coords <- session$userData$selected_pool
    
# Validate coordinates to ensure they are within Zurich area
    if (from_coords[1] < 8 || from_coords[1] > 9 || from_coords[2] < 47 || from_coords[2] > 48 ||
        to_coords[1] < 8 || to_coords[1] > 9 || to_coords[2] < 47 || to_coords[2] > 48) {
      showNotification("Coordinates are out of expected range.", type = "error")
      return()
    }
    
# Retrieve route from OpenRouteService using the selected transport type
    route <- get_route(from_coords, to_coords, input$transport_type)
    
    if (is.null(route)) {
      showNotification("Route not found.", type = "error")
      return()
    }
    
    leafletProxy("map") %>% 
      clearGroup("route") %>% 
      addGeoJSON(route, color = "blue", weight = 5, opacity = 0.7, group = "route")  
    showNotification("Route angezeigt.", type = "message")
  })
  
# Clearing routes
  observeEvent(input$clear_routes, {
    leafletProxy("map") %>% 
      clearGroup("route")
    showNotification("Routen gelöscht.", type = "message")
  })
  
# Printing latitude, longitude, and geolocation info
  output$lat <- renderPrint({
    req(input$user_lat)  
    input$user_lat
  })
  
  output$long <- renderPrint({
    req(input$user_lon)
    input$user_lon
  })
  
  output$geolocation <- renderPrint({
    req(input$geolocation)
    input$geolocation
  })
  
  output$geo_error <- renderPrint({
    input$geolocation_error
  })
  
# Observe the 'Standort aktualisieren' button click
  observeEvent(input$update_location, {
    session$sendCustomMessage(type = 'getLocation', message = 'update')
    showNotification("Standort aktualisiert", type = "message")
  })
  
# Custom message handler for initiating geolocation
  tags$script(HTML("
Shiny.addCustomMessageHandler('getLocation', function(message) {
  if (navigator.geolocation) {
    navigator.geolocation.getCurrentPosition(showPosition, showError, {enableHighAccuracy: true, timeout: 5000, maximumAge: 0});
  } else {
    Shiny.setInputValue('geolocation_error', 'Geolocation is not supported by this browser.');
  }
});
"))
  
# Updating the title select input
  observe({
    sorted_titles <- sort(unique(df_pools$Title))
    updateSelectInput(session, "title", choices = c("Alle", sorted_titles))
  })
  
# Reactive expression for filtered data
  map_df = reactive({
    temp_filtered <- df_pools %>%
      filter(Wassertemperatur >= input$temperatur[1], Wassertemperatur <= input$temperatur[2])
    
    status_filtered <- if (input$status != "Alle") {
      temp_filtered %>%
        filter(Status == input$status)
    } else {
      temp_filtered
    }
    
    title_filtered <- if (input$title != "Alle") {
      status_filtered %>%
        filter(Title == input$title)
    } else {
      status_filtered
    }
    
    title_filtered %>%
      filter(!is.na(Longitude), !is.na(Latitude)) %>%
      st_as_sf(coords = c("Longitude", "Latitude")) %>%
      st_set_crs(4326)
  })
  
# Initial map rendering
  output$map <- renderLeaflet({
    leaflet() %>%
      addTiles() %>%
      setView(lng = 8.5417, lat = 47.3769, zoom = 12) 
  })
  
# Updating the map with user location and pool markers
  observe({
    leafletProxy("map", data = NULL) %>%
      clearMarkers()
    
# Add user location marker if available
    if (!is.null(input$user_lat) && !is.null(input$user_lon)) {
      leafletProxy("map") %>%
        addMarkers(lng = input$user_lon, lat = input$user_lat, popup = "Your location")
    }
    
# Add pool markers
    leafletProxy("map") %>%
      addCircleMarkers(
        data = map_df(),
        popup = ~paste("<b>Bad:</b>", Title, "<br><strong>Wassertemperatur:</strong>", Wassertemperatur, "°C",
                       "<br><b>Status</b>", Status, "<br>Zuletzt aktualisiert:", Update),
        radius = 8,
        color = '#007BFF',
        fillOpacity = 0.7
      )
  })
}

#Launching the app by connecting the defined user interface (ui) and server logic (server) components

shinyApp(ui, server)
```

## Result (App Features)
The Zurich Public Pools App described in the code above provides the following features:

* **Real-Time Data Integration**: Access up-to-date water temperatures and the opening status of each public outdoor pool.
* **Interactive Map**: Navigate through an interactive map displaying all the pools. Click on any pool marker to get detailed information.
* **Route Planning**: Get directions to any selected pool from your current location by choosing your preferred mode of transportation (walking, cycling, or driving).
* **Customizable Searches**: Filter pools based on water temperature or opening status.
